\mytitle

### 2012-03-23 papers/SAS2012/implicitization.tex
Suppression de quelques vspace négatifs
Réduction des annexes aux tableaux.

\begin{figure}
  \vspace{-1em}
  \begin{minipage}{.3\linewidth}
    \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
  \end{minipage}
\hfill
\begin{minipage}{.5\linewidth}
\includegraphics[scale=0.7]{gopan_reps}
\vspace{-1em}
\begin{equation*}
y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
\end{equation*}
\end{minipage}
\caption{Example program and its invariant: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

@@

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i$ the abstract value
associated to $p_i$ :
\vspace{-.5em}
\[
f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
 \neg X_j)
\vspace{-.1em}
\]
The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.
Each time the invariant candidate of a point $p_j$ has been updated, $p_j$ is
inserted into $A'$ since it may be the start of a new feasible paths.

@@

This new subgraph contains all the paths from the previous one, and also new
paths that become feasible regarding the current abstract values.
The new paths in $P$ are computed one after another, until no more path
can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
Algorithm~\ref{algo:combined}, which corresponds to
Algorithm~\ref{algo:computepaths}. We also use SMT solving to discover
these new paths, but we subtly change the SMT formula given to the
SMT solver: we now try to find a path that is not yet in $P$, but is feasible
and makes the invariant candidate of its destination grow.
% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
We thus check the satisfiability of the formula
$f'(p_i)$, where:
\vspace{-.5em}
\[
f'(p_i) = f(p_i)\wedge \neg P
\vspace{-1em}
\]
$X_j$ is updated using an abstract union when the point $p_j$ is the target of a new path. This way, further SMT queries do not
compute other paths with the same source and destination if it is not needed
(because these new paths would not make $X_j$ grow, hence would not be returned
by the SMT solver).

@@

\subsection{Analysis Algorithm}
\label{sec:analysis-algorithm}
For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$. Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \cite{BourdonclePhd}.

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
\avirer{LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.}

SSA introduces $\phi$-functions at the head of a control code to
define variables whose value depends on which incoming edge to this
control node was last taken. For instance, for
\lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as
$\phi(2x_1+1,0)$. In this framework, variables are either defined as
an arithmetic ($+$, $-$, $\times$, $/$) function of other variables,
or not be representable as affine linear functions, because they are
defined using $\phi$-functions, bitwise arithmetic, loads from memory,
or return values from function calls.

This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g. as dimensions in polyhedra). A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested previously \cite{Colby_PhD,DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge. In this case, our tool just propagates symbolic values through the predecessor node, without introducing $\phi$-variables.

We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.

Our tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers. We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
The remaining memory accesses are treated as undefined values, as are the return
values of function calls. For better precision, we also apply function inlining
to the program, and we unroll every loop once.

Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers. Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g. the Astr\'ee system \cite{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.

Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \cite{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.

\subsection{Precision of the Various Techniques}
\label{sec:compare_techniques}

\MM{TODO: graphique supprimé ici}

@@

For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a
strictly stronger invariant, denoted by $\subsetneq$ (resp. $\supsetneq$),
and the proportion of control points where the invariants given by $T_1$ and
$T_2$ are uncomparable for the inclusion ordering (the remainder of the
control points are thus those for which both techniques give the same
invariant). We use convex polyhedra as the abstract domain.

Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
\emph{Guided Static Analysis} from \cite{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.21\%$ of the control points in $P_R$.
\emph{Path-focusing} from \cite{Monniaux_Gonnord_SAS11} gives
statistically better results, and finds better invariants than \emph{Guided
Static Analysis} in $5.00\%$ of the cases. However, it also loses
precision compared to \emph{Guided Static Analysis}  in an
important number ($4.00\%$) of control points.

Unfortunately, these two techniques are less precise than the classical analysis
in a important number of cases ($4.64\%$ for $G$, $5.14\%$ for $PF$).
This could be explained by the non-monotonicity of the results with widening operators.

Finally, our combined technique gives the most promising results, since it is
statistically more precise than the other techniques.

The analysis using disjunctive invariants greatly improves the
precision of the analysis (for $14.46\%$ of the control points in $P_R$), at an
acceptable time cost compared to our combined algorithm (see Table \ref{tab:time}). \MM{TODO}

While experiencing with techniques that use SMT-solving, we encountered some
limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
the SMT-solver is not able to decide the satisfiability of some SMT-formulae
expressing the semantics of non-linear programs. 
In this case, we skipped the functions for which the SMT-solver returned the
``unknown'' result.
This limitation occurred very rarely in our experiments, except for the analysis
of \emph{Lapack/Blas}, where 668 over the 1498 functions have been skipped.
\emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.

In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
(i) \emph{Linearization}, as per \cite{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
(ii) Replacing the results of nonlinear operations by ``unknown''.
Neither is currently implemented in our tool.

Table \ref{tab:time} \MM{TODO} gives the execution time of the different analysis
techniques. It is interesting to see that 
\emph{Path-focusing} technique is sometimes faster than the classical algorithm.
This seems due to the fact that this technique avoids many convex hulls, which
is costly, and also computes the image of an abstract value by an entire path
instead of many single transitions, thus leading to a large parallel assignment.
The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.

\subsection{Precision of Abstract Domains}
\label{sec:compare_domains}
\begin{sidewaystable}
\begin{center}
\setlength{\tabcolsep}{0.75ex}
\begin{tabular}{|l
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{PK/OCT}}
& \multicolumn{3}{c|}{\textbf{PK/BOX}}
& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
& \multicolumn{3}{c|}{\textbf{PK/GRID}}
& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table_domain.tex}
\end{tabular}
\end{center}
\caption{Results of the comparison of the various abstract domains, when using
the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
POLY),
Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
(GRID). The last column compares the domain of Convex Polyhedra with the
improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
using the classical widening operator (POLY). POLY and POLY* use the Parma
Polyhedral Library \cite{PPL}.
$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
\label{tab:domain}
\end{sidewaystable}

For each program and each pair $(D_1,D_2)$ of abstract domains, we compare by
inclusion the invariants of the different control points in $P_R$ (where $P_R$
is defined as in \ref{sec:compare_techniques}).
The main results are given in Table \ref{tab:domain}.

Statistically, the domain of convex polyhedra gives the better results, but
commonly yields weaker invariants than the domains of octagons/intervals;
this is a known weakness of its widening operator~\cite{Monniaux_LeGuen2011}. 
The Octagon domain appears to be much better than intervals; this is
unsurprising since in most programs and libraries, bounds on loop indices are
non constant: they depend on some parameters (array sizes etc.).

The Lapack/Blas benchmarks are unusual compared to the other programs. These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.

The analysis of linear equalities (PKEQ) performs very fast compared
to other abstract domains, but yields very imprecise invariants:
it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
The comparison between the domain of convex polyhedra and linear congruences
shows the percentage of points where the use of congruences refines the
invariant. This is the case for $1.04\%$ of the control points in $P_R$.

Finally, we evaluated the benefits of the improved version of the widening
operator for convex polyhedra from \cite{BagnaraHRZ05SCP}, compared to the
classical widening. We found that the improved version from
\cite{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
control points in $P_R$.

>>>>

\begin{figure}
%  \vspace{-1em}
  \begin{minipage}{.3\linewidth}
    \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
  \end{minipage}
\hfill
\begin{minipage}{.5\linewidth}
\includegraphics[scale=0.7]{gopan_reps}
%\vspace{-1em}
\begin{equation*}
y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
\end{equation*}
\end{minipage}
\caption{Example program and its invariant: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

@@

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i$ the abstract value
associated to $p_i$ :
%\vspace{-.5em}
\[
f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
 \neg X_j)
%\vspace{-.1em}
\]
The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.
Each time the invariant candidate of a point $p_j$ has been updated, $p_j$ is
inserted into $A'$ since it may be the start of a new feasible paths.

@@

This new subgraph contains all the paths from the previous one, and also new
paths that become feasible regarding the current abstract values.
The new paths in $P$ are computed one after another, until no more path
can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
Algorithm~\ref{algo:combined}, which corresponds to
Algorithm~\ref{algo:computepaths}. We also use SMT solving to discover
these new paths, but we subtly change the SMT formula given to the
SMT solver: we now try to find a path that is not yet in $P$, but is feasible
and makes the invariant candidate of its destination grow.
% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
We thus check the satisfiability of the formula
$f'(p_i)$, where:
%\vspace{-.5em}
\[
f'(p_i) = f(p_i)\wedge \neg P
%\vspace{-1em}
\]
$X_j$ is updated using an abstract union when the point $p_j$ is the target of a new path. This way, further SMT queries do not
compute other paths with the same source and destination if it is not needed
(because these new paths would not make $X_j$ grow, hence would not be returned
by the SMT solver).

@@

Experimental results of our tool on the benchmark presented in
Fig~\ref{fig:projects} are given below. The precision of different
techniques is given in Table~\ref{tab:techniques}. The precision of
abstract domains and widening operators in given in
Table~\ref{tab:domain}.

@@

\label{sec:compare_domains}
\begin{sidewaystable}
\begin{center}
\setlength{\tabcolsep}{0.75ex}
\begin{tabular}{|l
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{PK/OCT}}
& \multicolumn{3}{c|}{\textbf{PK/BOX}}
& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
& \multicolumn{3}{c|}{\textbf{PK/GRID}}
& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table_domain.tex}
\end{tabular}
\end{center}
\caption{Results of the comparison of the various abstract domains, when using
the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
POLY),
Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
(GRID). The last column compares the domain of Convex Polyhedra with the
improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
using the classical widening operator (POLY). POLY and POLY* use the Parma
Polyhedral Library \cite{PPL}.
$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
\label{tab:domain}
\end{sidewaystable}
